---
title: "Single cell RNA-seq in R and python"
title-short: "CDNM workshop sketch"
format: html
toc: true
editor_options: 
  chunk_output_type: console
---

# Introduction

This document is written in markdown for processing
with quarto.

## Using R/Bioconductor to get the TENx PBMC 3k dataset
```{r, message=FALSE}
library(TENxPBMCData)
library(SingleCellExperiment)
p3k = TENxPBMCData("pbmc3k")
p3k
class(assay(p3k))
```

## Using python/scanpy to get the TENx PBMC 3k dataset

```{python, message=FALSE}
import scanpy as sc
pp3k = sc.datasets.pbmc3k()
pp3k
```

# Bioconductor 'SingleR' workflow for reference-based labeling

Start with the single-cell expression data for 2700
PBMCs.  Raw counts are normalized and log-transformed,
see `scuttle::normalizeCounts`.

```{r getp3, message=FALSE}
assay(p3k) = as.matrix(assay(p3k)) # make dense
p3k = scuttle::logNormCounts(p3k)
p3k
```

Develop a provisional labeling with SingleR.


Here we use HumanPrimaryCellAtlas "fine-grained"
labeling.

```{r dosing, message=FALSE,eval=TRUE,cache=TRUE}
library(SingleR)
library(celldex)
hpca = celldex::HumanPrimaryCellAtlasData()
inirown = rownames(p3k)
rownames(p3k) = make.names(rowData(p3k)$Symbol, unique=TRUE)
ann2 = SingleR(p3k, hpca, labels=hpca$label.fine)
library(scater)
p3kp = runPCA(p3k)
pairs(reducedDims(p3kp)$PCA[,1:4],
  col=factor(ann2$labels), pch=19, cex=.3)
rownames(p3k) = inirown
```

Interactive visualization

```{r, doviz, message=FALSE}
df23 = data.frame(reducedDims(p3kp)$PCA[,1:4], type=ann2$labels)
library(ggplot2)
p23 = ggplot(df23, aes(PC3, PC2, color=type)) + geom_point()
library(plotly)
ggplotly(p23)
```
